R Notebook

### 2017-03-13 manuscript/draft.Rmd
Dave's edits

* BBS + retriever
* Monthly climate data from interpolated weather stations (PRISM)
* NDVI (GIMMS)
* Altitude (SRTM 90m Digital Elevation Database via raster::getData)

### Accounting for site \& observer effects
* Observer effects are inherent in
  large data sets taken by different observers, and are known to occur in BBS 
  [citation].
* Mixed model for observer effort \& site effects: used Stan, included random
  effects observer_id to account for situations were one observer consistently
  found more (or fewer) species than average. These models also included a 
  site-level random effect to account for differences in actual richness across 
  BBS routes. [[Maybe year effect?]].
  No fixed effects because not interested in explaining differences among sites
  at this point; just want to account for their existence.
* Residuals from this model give an estimate of how richness changed over time
  at each site, accounting for observer differences.  We then used this "corrected"
  version of the data for our time-series models such as the AR1, etc.
* Note: throughout this manuscript, we use Gaussian errors for richness rather than 
  a proper count distribution because:
  * underdispersion (variance/mean << 1)
  * Easier to access out-of-the-box time series methods like auto-ARIMA.
  * Central limit theorem, large number of species (mean ~50 spp)

@@

**Baseline models**
* The lme4 model fitted above.  This model assumes that the site-level average
  is constant, and tries to estimate its value.  The confidence intervals don't
  expand over time (except when observers change) because the model says nothing
  has changed
* Naive AR1 model (adjusted for observer effects). This model says that richness 
  can diffuse up or down from the previous year's value, without any constraints.
  As a result, the confidence intervals expand quickly over time.

**Time series models**
* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
  model selection based on AIC.
* Auto-ARIMA + environment.

**SDMs**
* SDMs: use graf model w/ mixed model estimates of site \& observer effects
  as predictors (include error in predictors as estimated from the mixed model
  to propagate our uncertainty forward). This is basically SDM, but with a 
  better model (graf) and more information (monthly climate data instead of 
  using constant values from the Worldclim data, and observer effects). We used 
  Poisson-binomial confidence intervals, as discussed in Calabrese et al. 2014.
* JSDM. Should get best of the previous 2. Might be able to propagate uncertainty
  from lme4.

**"Macroecological" model**
* richness-level GBM w/ mixed model point estimates as bonus predictors where 
  available. This lets us predict richness directly instead of via noisy 
  species-level estimates, but doesn't use the full data set in one model.
  Maybe need to run multiple times to propagate uncertainty from lme4.

@@

Evaluated each year after the last training year (2010?):
* (R)MSE
* Log-likelihood (simultaneously rewards good point estimates, precision, and coverage)
* Coverage at 95%
* Calibration/bias

@@

>>>>

* BBS + retriever
* Predictor variables
  * Monthly climate data from interpolated weather stations (PRISM), converted
    to BIOCLIM variables. Many of these variables were tightly correlated, so
    we only included the 8 variables selected by Harris's (2015) BBS analysis
  * NDVI (GIMMS)
  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
* Data summary: 
  * Number of sites, species, unique observers
  * central moments of richness (mean about 51, sd about 12 spp)
  * The within-site variation was much smaller (sd about 5 spp), making the
    site-level distributions underdispersed.  
    * This underdispersion implies
      that common count models such as binomial, Poisson, negative binomial 
      won't fit well.
    * For this reason (and because the time series methods assumed 
      Gaussian errors), we defined model performance throughout this paper in 
      terms of Gaussians (e.g. mean squared error), rather than with count
      distributions.
        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
          distributions internally, however.

### Accounting for site \& observer effects
* Observer effects are inherent in large data sets taken by different 
  observers, and are known to occur in BBS [citation].
* While these biases are known to exist, they are difficult to quantify: even if
  the different observers ran the same transects under identical conditions (which 
  they don't), we would still only have an estimate of their differences, not an 
  exact value.
* We estimated the observer effects, along with our uncertainty about those
  effects' size & direction, with a linear mixed model.
* This model partitions the variance in observed richness values into site-level
  variance, observer-level variance, and residual variance (e.g. variation within
  a site from year to year or from morning to morning).
  * No fixed effects in this model (e.g. environmental predictor variables because 
    we're not interested in explaining differences among sites at this point; we 
    just want to account for their existence while estimating observer effects.
  * Subtracting off the estimated observer effect provides an estimate of how many
    would have been found by a "typical" observer on that day; in this sense, it 
    enables us to correct for observer differences. We then used this "corrected" 
    version of the data for our time-series models such as the AR1, etc.
* A simple model like this one can't tell us exactly how each observer differed from
  the others, so it is important to represent our uncertainty about these differences.
  Here, we did so by collecting [[N]] Monte Carlo samples from the model's posterior
  distribution, using Stan. This provided us with [[N]] plausible values for each 
  observer, so we did not have to commit to a single value.
* While some methods for predicting richness might be able to use the uncertain
  estimates of the observer effects directly, we take the simpler approach of just
  fitting each of the downstream models with each of the Monte Carlo samples. This
  increases the amount of computational work by a factor of [[N]], but CPU time
  wasn't our limiting resource.

@@

**Baseline models**
* The lme4 model fitted above.  This model assumes that the site-level average
  is constant, and tries to estimate its value.  The confidence intervals don't
  expand over time (except when observers change) because the model says nothing
  has changed.
* Naive AR1 model (fitted to individual sites' residuals from the observer model). 
  This model says that richness can diffuse up or down from the previous year's 
  value, without any constraints. As a result, the confidence intervals expand 
  quickly over time.
* Fit once per MCMC sample, to account for uncertainty in what observer effect
  values we should subtract off

**Time series models**
* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
  model selection based on AIC.
* Auto-ARIMA + environment.
* Uncertainty treated as above

**"Macroecological" model**
* Use a nonlinear model to predict richness as a function of environmental predictors
  (BIOCLIM + elevation + NDVI).
* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
  noisy species-level estimates, but doesn't use the full data set in one model.
* For this part of the analysis, we used the `gbm` package for fitting boosted regression
  trees with Gaussian errors.  
* We included observer effects & the associated uncertainty by running this model [[N]]
  times (once per MCMC sample).
* We did *not* include the site-level random effects as predictors, meaning that this
  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
  NDVI values should have identical richness distributions.

**SDMs**
* Probably the most common approach?
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we
  used a custom wrapper around the randomForest package. Random forests are flexible
  models that are built by combining many independent models ("trees") fit to different 
  versions of the training data, which made it straightforward to adapt them to our 
  situation: we simply fit one tree using each of the Monte Carlo estimates of our 
  observer effects and averaged the predictions of the resulting forest.

**JSDM** 
* New approach that tries to get the , cite examples.
* Should get best of the previous 2. Might be able to propagate uncertainty
  from lme4.

@@

Evaluated each year after the last training year (2010?):
* (R)MSE 
  * How far, on average, are the models' predictions from the true value?
  * [[RMSE has better units than MSE (species versus squared species), but MSE is
    nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
* Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
  and coverage)
* Coverage at 95%
* Calibration/bias

@@

### 10. [[?]]

Sites' biotic and abiotic environments differ in more ways than the 
few ones you happened to measure. As a result, some sites will 
consistently have more species than a regression model (or sum of 
regression models) would predict, while others willl consistently have 
fewer. Depending on the spatial scale of these unmeasured variables, 
independent site-level random effects could be sufficient for dealing 
with this, or spatially-autocorrelated random effects might be needed.
